Chasing \rainbows and \oceanglint:\\ Inner working angle constraints for the Habitable Worlds Observatory

### 2023-05-12 src/tex/ms.tex
[showyourwork] overleaf sync

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\HWO), a coronagraphic space mission to detect rocky planets in habitable zones, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase curves of starlight reflected by an exoplanet can reveal ocean glint, rainbows and other phenomena caused by scattering by clouds or atmospheric gas.
%
Direct imaging missions are optimised for planets near quadrature, and \HWO' coronagraph may obscure the phase angles where such optical features are strongest. 
%
The range of accessible phase angles for a given exoplanet will depend on the planet's orbital inclination and/or the coronagraph's inner working angle (IWA). 
%
For planets in near edge-on orbits, accessible phase angles are limited by coronagraphic obscuration. 
%
We use \HWO'~catalogue of 160 target stars to estimate the number of exo-Earths that could be searched for ocean glint, rainbows, and Rayleigh scattering. 
%
We find that any exo-Earths in planetary systems in this catalogue will have an accessible Rayleigh scattering peak in their polarimetric phase curve. 
%
The rainbow due to water clouds at scattering angles of \qtyrange{140}{160}{\degree} would be accessible to \HWO\ in $\sim\num{42}$ systems, while ocean glint signature at scattering angles of $\qtyrange{50}{70}{\degree}$ would be accessible in $\sim\num{14}$ systems, assuming an \IWA of \qty{62}{\mas} ($3\,\lambda/D$ for a $D=\qty{6}{\meter}$ telescope at $\lambda=\qty{600}{\nano\meter}$).
%
Improving the \IWA to $2\,\lambda/D$ increases the number of systems with accessible rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases slightly for eccentric orbits.  %and scales approximately inversely with the \IWA.  
\end{abstract}

@@

In general, the glint on a planet will be strongest at red wavelengths because there atmospheric Rayleigh scattering and attenuation are smallest \citep{Zugger_2011}. This has indeed been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}. %A degeneracy between polar brightness and glint is also smaller at redder wavelengths \citep{2012ApJ...752L...3C}.
%
%The width across phase angles of the glint feature in total and polarized flux generally broadens with increasing wave height \citep{2008Icar..195..927W, kopparla2018, treesandstam2019, trees2022}, which depends on the surface wind speed, as described by \citet{CoxMunk1954}.
%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
    \centering
    \includegraphics{figures/figure-1-bott-plot.pdf}
    \script{create-figure-1-bott-plot.py}
    \caption{
        The phase curves of the normalized linearly polarized flux of a cloudy planet with a wavy ocean surface and an Earth-like atmosphere in an edge-on orbit; adapted from \citet{treesandstam2019}.  
        %(the planet goes through all possible phase angles throughout its orbit). 
        The wind speed over the ocean, which determines
        the wave height \citep[][]{CoxMunk1954}, is \qty{7}{\meter\per\second}. 
        The clouds consist of spherical water droplets. 
        The cloud coverage fraction is 0.5 and while the  
        cloud pattern is patchy, the glint is always cloud-free. 
        %to maximize the glint signature \citep[see also Fig. 9 of][from which we took the mean signal of 300 planets with randomly generated patchy cloud patterns + 2 standard deviations, representative for cloud-free glint scenarios; the \enquote{noise} at the larger phase angles and longer wavelengths is due to the variation in the location of cloud patches]{treesandstam2019}. 
        %The polarized flux is normalized to the total (unpolarized + polarized) flux (it is thus in fact the degree of polarization). 
        % Note that the reference plane for the polarized flux is chosen such that only the polarized components perpendicular or parallel to the plane differ significantly from zero.
        % and the latter
        The fluxes are normalized such that the total (unpolarized + polarized) flux equals the planet's geometric albedo at phase angle $\alpha = 0^\circ$.
        % The cloudy scenario includes a low-altitude, Earth-like water 
        % cloud deck. 
        Below the plot, we show the scattering angles of glories, rainbows, Rayleigh scattering, and ocean glint. Note that the features in the actual curves partly overlap with each other.
        %Throughout the paper we approximate feature sensitivity ranges as those that capture the rise and fall of peaks in polarized light for most habitable zone species in the UV, Optical, and IR (phase angles: glint $\sim\qtyrange{50}{70}{\degree}$; rainbow $\sim\SIrange{140}{160}{\degree}$; Rayleigh narrow of quadrature, and other forward and backscattering effects at the widest and narrowest phase angles).
    }
    \label{fig:bottplot}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Rainbows, cloudbows, glories and other scattering phenomena can dramatically influence the shape of a planet's photometric and polarimetric phase curves and provide a wealth of information about exoplanets \citep{karalidi2012rainbow, stam2008,bailey2007,2014A&A...566L...1G}.%
\footnote{
    Rainbows occur when light is scattered by rain droplets that are large compared to the wavelength of the light. 
    Cloudbows refer to features occurring at the same phase angles from smaller cloud particles, with the spread in colours invisible to the human eye making them appear white \citep[for further discussion, see][]{bailey2007}.  
}
%
In the Solar System, the polarized back-scatter from Titan indicated that the moon's reflection was due to an atmospheric haze rather than a surface \citep{zellner1973polarization}, Venus's polarized phase curve at three wavelengths showed that the clouds were $\sim\qty{75}{\percent}$ H\textsubscript{2}SO\textsubscript{4}--water mixture and constrained the droplet size distribution \citep{hansenhovenier1974}.
%
The glory feature in Venus's flux phase curve was later used to rule out various species as the unknown UV absorber \citep{petrova2018glory}.
%In unpolarized light, the scattering by cloud particles can dramatically change the shape of a planet's phase curve compared to that of a Lambertian reflecting planet.
%
Glories and rainbows are technically visible both in total and polarized flux phase curves in the presence of spherical cloud and/or haze particles, but their strongly polarized nature coupled with the suppression of other polarized features at gibbous and full phases make them more easily characterized with polarimetry than with photometry \citep{karalidi2011, stam2008, treesandstam2019,trees2022}.
%

@@

\Cref{fig:accessible_phase_angles} and Table \ref{tab:nstars_detect} shows the number of systems in which key scattering phenomena will be observable for different \IWA.
%The \IWA of \HWO's coronagraph will depend on the primary mirror diameter, observing wavelength, and how well a coronagraph of a given size can suppress the glare of the host star.
%The smaller the \IWA, the more features we will be observable and in more systems. 
The Rayleigh feature, which can be used to determine if a planet has an atmosphere, appears near quadrature and is therefore observable in the most systems.
The rainbow feature appears further from quadrature (around $\alpha=\qty{40}{\degree}$), but will still be observable in a significant fraction of systems.
The position and shape of the rainbow peak is sensitive to the properties of the droplets, allowing their nature to be determined if the peak of the rainbow feature can be reached.
The ocean glint feature occurs even further from quadrature, and so would be observable in fewer systems. 
In most systems, only the first half of the glint feature would be observable, but this can still be used to identify a liquid ocean on the surface of a planet, a key indicator of habitability. 
%Finally, it is possible to view the glory scattering feature in one system, Alpha~Cen~A.

@@

SRV and JLB acknowledge funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation program under grant agreement n°805445.
SLC acknowledges support from an STFC Ernest Rutherford Fellowship. 
TDG acknowledges funding from the Max Planck ETH Center for Learning Systems.
KMB acknowledges support from NASA Habitable Worlds grant n°80NSSC20K152, and previous support for related work from NASA Astrobiology Institute's Virtual Planetary Laboratory under Cooperative Agreement n°NNA13AA93A.
IL acknowledges the support by a postdoctoral grant issued by the Centre National d'Études Spatiales (CNES) in France.
PB, IL, and YG were supported by the Action Spécifique Haute Résolution Angulaire (ASHRA) of CNRS/INSU co-funded by CNES.
EHP is supported by the NASA Hubble Fellowship grant \#HST-HF2-51467.001-A awarded by the Space Telescope Science Institute, which is operated by the Association of Universities for Research in Astronomy, Incorporated, under NASA contract NAS5-26555.
LA and EC acknowledge funding from the European Research Council (ERC) under the European Union's Horizon Europe research and innovation programme (ESCAPE, grant agreement n°101044152).
OA and LK acknowledge funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement n°819155)
SYH was funded by the generous support of the Heising-Simons Foundation.
RJP is supported by NASA under award number 80GSFC21M0002.
OHS acknowledges funding from the direction scientifique générale de l'ONERA (ARE Alioth).

>>>>

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (\HWO), a coronagraphic space mission to detect rocky planets in habitable zones, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase curves of starlight reflected by an exoplanet can reveal ocean glint, rainbows and other phenomena caused by scattering by clouds or atmospheric gas.
%
Direct imaging missions are optimised for planets near quadrature, and \HWO' coronagraph may obscure the phase angles where such optical features are strongest. 
%
The range of accessible phase angles for a given exoplanet will depend on the planet's orbital inclination and/or the coronagraph's inner working angle (IWA). 
%
For planets in near edge-on orbits, accessible phase angles are limited by coronagraphic obscuration. 
%
We use \HWO'~catalogue of 160 target stars to estimate the number of exo-Earths that could be searched for ocean glint, rainbows, and Rayleigh scattering. 
%
We find that any exo-Earths in planetary systems in this catalogue will have an accessible Rayleigh scattering peak in their polarimetric phase curve. 
%
The rainbow due to water clouds at scattering angles of \qtyrange{140}{160}{\degree} would be accessible to \HWO in $\sim\num{42}$ systems, while ocean glint signature at scattering angles of $\qtyrange{50}{70}{\degree}$ would be accessible in $\sim\num{14}$ systems, assuming an \IWA of \qty{62}{\mas} ($3\,\lambda/D$ for a $D=\qty{6}{\meter}$ telescope at $\lambda=\qty{600}{\nano\meter}$).
%
Improving the \IWA to $2\,\lambda/D$ increases the number of systems with accessible rainbows and glints by factors of roughly 2 and 3, respectively.
%
The number of systems in which an exo-Earth could be searched for these scattering phenomena increases slightly for eccentric orbits.  %and scales approximately inversely with the \IWA.  
\end{abstract}

@@

In general, the glint on a planet will be strongest at red wavelengths because there atmospheric Rayleigh scattering and attenuation are smallest \citep{Zugger_2011}. This has been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}. %A degeneracy between polar brightness and glint is also smaller at redder wavelengths \citep{2012ApJ...752L...3C}.
%
%The width across phase angles of the glint feature in total and polarized flux generally broadens with increasing wave height \citep{2008Icar..195..927W, kopparla2018, treesandstam2019, trees2022}, which depends on the surface wind speed, as described by \citet{CoxMunk1954}.
%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
    \centering
    \includegraphics{figures/figure-1-bott-plot.pdf}
    \script{create-figure-1-bott-plot.py}
    \caption{
        The phase curves of the normalized linearly polarized flux of a cloudy planet with a wavy ocean surface and an Earth-like atmosphere in an edge-on orbit; adapted from \citet{treesandstam2019}.  
        %(the planet goes through all possible phase angles throughout its orbit). 
        The wind speed over the ocean, which determines
        the wave height \citep[][]{CoxMunk1954}, is \qty{7}{\meter\per\second}. 
        The clouds consist of spherical water droplets. 
        The cloud coverage fraction is 0.5 and while the  
        cloud pattern is patchy, the glint is always cloud-free. 
        %to maximize the glint signature \citep[see also Fig. 9 of][from which we took the mean signal of 300 planets with randomly generated patchy cloud patterns + 2 standard deviations, representative for cloud-free glint scenarios; the \enquote{noise} at the larger phase angles and longer wavelengths is due to the variation in the location of cloud patches]{treesandstam2019}. 
        %The polarized flux is normalized to the total (unpolarized + polarized) flux (it is thus in fact the degree of polarization). 
        % Note that the reference plane for the polarized flux is chosen such that only the polarized components perpendicular or parallel to the plane differ significantly from zero.
        % and the latter
        The fluxes are normalized such that the total (unpolarized + polarized) flux equals the planet's geometric albedo at phase angle $\alpha = 0^\circ$, scattering angle $\varphi = 180^\circ$.
        % The cloudy scenario includes a low-altitude, Earth-like water 
        % cloud deck. 
        Below the plot, we show the scattering angles of glories, rainbows, Rayleigh scattering, and ocean glint. Note that the features in the actual curves partly overlap with each other.
        %Throughout the paper we approximate feature sensitivity ranges as those that capture the rise and fall of peaks in polarized light for most habitable zone species in the UV, Optical, and IR (phase angles: glint $\sim\qtyrange{50}{70}{\degree}$; rainbow $\sim\SIrange{140}{160}{\degree}$; Rayleigh narrow of quadrature, and other forward and backscattering effects at the widest and narrowest phase angles).
    }
    \label{fig:bottplot}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Rainbows, cloudbows, glories and other scattering phenomena can dramatically influence the shape of a planet's photometric and polarimetric phase curves and provide a wealth of information about exoplanets \citep{karalidi2012rainbow, stam2008,bailey2007,2014A&A...566L...1G}.% Detailed characterization of the rainbow in phase and wavelength space can constrain the condensate species, providing context to the habitability of the planet.
\footnote{
    Rainbows occur when light is scattered by rain droplets that are large compared to the wavelength of the light. 
    Cloudbows refer to features occurring at the same phase angles from smaller cloud particles, with the spread in colours invisible to the human eye making them appear white \citep[for further discussion, see][]{bailey2007}.  
}
%
In the Solar System, the polarized back-scatter from Titan indicated that the moon's reflection was due to an atmospheric haze rather than a surface \citep{zellner1973polarization}, Venus's polarized phase curve at three wavelengths showed that the clouds were $\sim\qty{75}{\percent}$ H\textsubscript{2}SO\textsubscript{4}--water mixture and constrained the droplet size distribution \citep{hansenhovenier1974}.
%
The glory feature in Venus's flux phase curve was later used to rule out various species as the unknown UV absorber \citep{petrova2018glory}.
%In unpolarized light, the scattering by cloud particles can dramatically change the shape of a planet's phase curve compared to that of a Lambertian reflecting planet.
%
Glories and rainbows are technically visible both in total and polarized flux phase curves in the presence of spherical cloud and/or haze particles, but their strongly polarized nature coupled with the suppression of other polarized features at gibbous and full phases make them more easily characterized with polarimetry than with photometry \citep{karalidi2011, stam2008, treesandstam2019,trees2022}.
\looseness=-1
%

@@

\Cref{fig:accessible_phase_angles} and Table \ref{tab:nstars_detect} show the number of systems in which key scattering phenomena will be observable for different \IWA.
%The \IWA of \HWO's coronagraph will depend on the primary mirror diameter, observing wavelength, and how well a coronagraph of a given size can suppress the glare of the host star.
%The smaller the \IWA, the more features we will be observable and in more systems. 
The Rayleigh feature, which can be used to determine if a planet has an atmosphere, appears near quadrature and is therefore observable in the most systems.
The rainbow feature appears further from quadrature (around $\alpha=\qty{40}{\degree}$), but will still be observable in a significant fraction of systems.
The position and shape of the rainbow peak is sensitive to the properties of the droplets, allowing their nature to be determined if the peak of the rainbow feature can be reached.
The ocean glint feature occurs even further from quadrature, and so would be observable in fewer systems. 
In most systems, only the first half of the glint feature would be observable, but this can still be used to identify a liquid ocean on the surface of a planet, a key indicator of habitability. 
%Finally, it is possible to view the glory scattering feature in one system, Alpha~Cen~A.

@@

SRV and JLB acknowledge funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation program under grant agreement n°805445.
SLC acknowledges support from an STFC Ernest Rutherford Fellowship. 
TDG acknowledges funding from the Max Planck ETH Center for Learning Systems.
KB acknowledges support from NASA Habitable Worlds grant n°80NSSC20K152, and previous support for related work from NASA Astrobiology Institute's Virtual Planetary Laboratory under Cooperative Agreement n°NNA13AA93A.
IL acknowledges the support by a postdoctoral grant issued by the Centre National d'Études Spatiales (CNES) in France.
PB, IL, and YG were supported by the Action Spécifique Haute Résolution Angulaire (ASHRA) of CNRS/INSU co-funded by CNES.
EHP is supported by the NASA Hubble Fellowship grant \#HST-HF2-51467.001-A awarded by the Space Telescope Science Institute, which is operated by the Association of Universities for Research in Astronomy, Incorporated, under NASA contract NAS5-26555.
LA and EC acknowledge funding from the European Research Council (ERC) under the European Union's Horizon Europe research and innovation programme (ESCAPE, grant agreement n°101044152).
OA and LK acknowledge funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement n°819155)
SYH was funded by the generous support of the Heising-Simons Foundation.
RJP is supported by NASA under award number 80GSFC21M0002.
OHS acknowledges funding from the direction scientifique générale de l'ONERA (ARE Alioth).
